---
title: 'Gallery: 3D stuff'
author: Knut Behrends
date: '2018-01-28'
slug: 3D
banner: img/gallery/thumb/rotated-cubes-50.png
categories: [fun]
tags: [creative, fun, 3D]
---


```{r echo=FALSE}
htmltools::HTML('<div class="alert alert-info"><b>TEST</b>
                There should be a 3D scene below

                </div>')
```

#### A red cube oribiting two other cubes

This 3D scene is from a Pluralsight course on [WebGL/three.js fundamentals](https://www.pluralsight.com/courses/webgl-threejs-fundamentals) by Alex Mackey. I have included this example here for testing purposes. I want to have a reference, or a simpler alternative, for verifying that [exporting 3D plots ](/post/2018/01/constructing-a-tiny-social-network-visualization-from-the-rstats-hashtag/) from the [R package $threejs$](https://bwlewis.github.io/rthreejs) works.

```{javascript eval=FALSE}
// Some javascript for debugging:
example.scene.getObjectByName("box");

```

```{r echo=FALSE}
htmltools::HTML('<div id="webgl-container"></div>
<script type="text/javascript" src="/js/vendor/dat.gui.min.js"></script>   
<script src="/js/vendor/three.min.js"></script>
<script src="/js/threejs-knbknb/threejs-orbiting-cube.js"></script>')
# //<script src="/js/threejs-knbknb/threejs-boilerplate.js"></script>
```

**Links to threejs stuff:**

- [Github Repo](https://github.com/mrdoob/three.js/)
- [Examples from Book: 3D Programming for Kids](https://pragprog.com/book/csjava/3d-game-programming-for-kids)
- [Examples from another Book](http://www.smartjava.org/content/all-109-examples-my-book-threejs-threejs-version-r63)
- [Three.js examples](http://stemkoski.github.io/Three.js/)
