---
title: Graphs (in the sense of networks)
draft: true
author: Knut Behrends
date: '2018-01-16'
slug: drawing-a-figshare-network-with-ggraph-and-tidygraph
categories:
  - R
  - networks
tags:
  - 'rstats'
  - 'tidyverse'
---

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
               echo = TRUE, tidy = FALSE, height = 3, width = 5,
               fig.width = 9)
options(digits = 3, tibble.print_min = 6)
options(knitr.table.format = "html") 

```

```{r intro,echo=FALSE}
htmltools::HTML('<div class="alert alert-success alert-info alert-warning alert-danger"><b>NOTE</b>
This blogpost is inspired by this one: <a href="https://nxskok.github.io/blog/2017/12/30/drawing-a-network-with-ggraph-and-tidygraph/">
Drawing a network with "ggraph" and "tidygraph" by Ken Butler.
</a><br>
</div>')
```



#### Example: the Figshare categories

Data provider Figshare models their contents with and elaborate hierarchy of disciplines.

I'm trying to learn the differences between using the *igraph* library directly and the new packages *ggraph* and *tidygraph*. 
Both new packages were authored by  [Thomas Lin Pedersen](https://github.com/thomasp85).
created two R packages: `tidygraph` for storing and handling graphs, and `ggraph` for
drawing them, `ggplot`-style. 

Load all needed packages:

```{r libs}
library(igraph)
library(httr)
library(tidyverse)
library(tidygraph)
library(ggraph)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set1")

```


### Drawing a network with "ggraph" and "tidygraph"


#### Reading in the data, and storage as a graph

The input file (part) is a JSON File.


```{r infile, echo=FALSE}

categories_file <- here::here("data_private/network/figshare-categories.RDS")

 if(! file.exists(categories_file)) {
        categories <- httr::GET("http://api.figshare.com/v2/categories") %>%
                content(as="text") %>%
                jsonlite::fromJSON(simplifyDataFrame = TRUE) %>% 
                  as_tibble() 
         saveRDS(categories, file = categories_file)
 } else {
         categories <- readRDS(categories_file)
 }

```

To read this in, I treat it as a three-column json-File. 
Quick look at the data:

```{r}
kable(head(categories %>% select(3,2,1), 50), 
      caption = "The Figshare Categories Hierarchy (first 50). Note that the parent IDs have not been assigned a name.", row.names = 1,
      align = "lll")
```

```{r echo=FALSE, eval=FALSE}
#View(categories %>% select(2,3,1) %>% dplyr::slice(1200:1300))
```

#### Quick visualization of how category IDs were assigned by Figshare Designers

Plot Code: 

```{r vis1}
p1 <- ggplot(categories, aes(id, I(1:nrow(categories)))) +
        geom_point(size=0.1) +
        xlab("Category ID") + ylab("Count") +
        ggtitle("Figshare Category IDs", 
                subtitle="Gaps indicate that a few category-IDs have been deleted or renamed")
p2 <-  ggplot(categories, aes(parent_id)) +
        geom_histogram(binwidth = 5) +
        xlab("Category Parent IDs") + 
        ylab("Count") + 
        ggtitle("Figshare Parent Category IDs", 
                subtitle="Gaps indicate that parent-category-IDs have small values or few distinct values > 1000")

```

The plot

```{r echo=FALSE}
gridExtra::grid.arrange(p1,p2, nrow=2, ncol=1)

```

Which category-ids are *not* assigned?

```{r ids, echo=FALSE}
unused_ids <- setdiff(1:max(categories$id), unique(categories$id))

data.frame(pos = 1:length(unused_ids), 
           id=unused_ids) %>% 
        ggplot(aes(pos, id)) +
        geom_point(size=0.2) + ylim(c(0,250)) +
        scale_x_continuous(limits=c(0,50)) +
        ggtitle("Unused Category IDs", 
                subtitle = "Roughly every 3rd number is not assigned to any category ID")

```



`tidygraph` has a function for turning a data frame structured like
this, with edges as rows, into an actual tidy graph:

```{r graph, echo=FALSE}

categories_g <- categories %>%
        select(id , parent_id) %>% 
        as_tbl_graph()
categories_g
```

How many attribute names are set?

```{r}
categories_g %>%
        vertex_attr() %>% 
        names()
```

Only one attribute "name" is set.

Which edges have the parent_id 48?

```{r eval=TRUE}

igraph::E(categories_g)[[inc('48')]]  %>%
        head(10)
```

There are `r length(igraph::E(categories_g)[[inc('48')]]  )` such edges.


```{r echo=FALSE}

categories_g <- set_edge_attr(categories_g,
                              "subject",
                              value=categories$title)
categories_g <- set_edge_attr(categories_g,
                              "id",
                              value=categories$id)
categories_g <- set_edge_attr(categories_g,
                              "parent_id",
                              value=categories$parent_id)
```

Now, how to assign names to the category ids.

The nodes and edges are shown separately, with the nodes numbered (those
numbers are used in the edges).

## Extracting the nodes and edges

`tidygraph` uses the same pipeline ideas as the `tidyverse`, but extra
care is required because a tidy graph object has both nodes and edges in
it. An extra verb `activate` lets you specify which one you would like
to be "primary" (but you can also access the other one). Piping the
output of `activate` into `as_tibble` lets you save the nodes or edges
as a data frame, separately:

```{r}

nodes = categories_g %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(node_number=row_number()) %>%
  print(n=10)
```

```{r}
edges = categories_g %>%
  activate(edges) %>%
  as_tibble() %>% print(n=10)
```

I saved the node numbers as well as names into the nodes data frame.
This lets me look up the *names* of the nodes referred to in the edges,
to convince myself that I have something sensible:

```{r}
edges %>%
  left_join(nodes,by=c("from"="node_number")) %>%
  left_join(nodes,by=c("to"="node_number")) %>%
  print(n=10)
```

Edges seem to have been determined properly by this operation. 
Now I can assign also The category names as attributes with the same method (TODO)

## Drawing the graph



Graphs can be plotted in many different ways. For this graph, I know
there is a representation in two dimensions, since the category hierarchy is quite complex.
In `ggraph`, node alignment in the plotting panel is controlled by the `layout` parameter of the `ggraph` command.  The *Kamada-Kawai* method tries to plot neighbouring nodes a constant distance
apart, and chooses their locations as if they are connected by
unit-length springs. 

Edges get drawn as lines and draw the nodes as points. I label the
nodes by which discipline they represent. 
These operations are accomplished by
`geom_edge_link`, `geom_node_point` and `geom_node_text`.

```{r echo=FALSE}
#  Additional
# notes: the `alpha=0.3` draws the edges partly transparent, so that the
#  names stand out; the use of `repel=T` "repels" the names
# away from the points they represent so that you can see them (in the
# manner of `geom_text_repel`):
```

```{r}
ggraph(categories_g,layout="kk")+
  geom_edge_link(colour="blue", alpha=0.1)+
  geom_node_point() +
        theme_graph()

 # geom_node_text(aes(label=id),size=2,repel=T)
```

This seems to do a nice job of reproducing the topology in a clear
fashion with no overlaps where there shouldn't be any. (I tried this
using the separate stations, Westgate and Kirkgate, in Wakefield, and
the display was a lot more confusing.) It doesn't get compass directions
right (it seems to have west and east reversed), but we had no right to
expect that it would.

#### Tabulating out-degrees


```{r}
g.outd <- degree(categories_g, mode = c("all"))
#names(g.outd) <- edge_attr(categories_g, "subject")
#View a summary of out-degree
table(g.outd)
```

