---
title: "UK NHS Medical Staff, 2009-2017"
author: "Knut Behrends"
date: "2018-01-23"
output: html_document
slug: uk-nhs-nurses
banner: img/post/thumb/uk-nhs-people1-50.png
categories:
  - society
tags:
  - 'rstats'
  - 'tidyverse'
  - medicine
  - society  
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
               echo = TRUE, tidy = FALSE, height = 3, width = 5,
               fig.width = 9)
options(digits = 3, tibble.print_min = 6)
options(knitr.table.format = "html") 
```

```{r intro,echo=FALSE}
htmltools::HTML('<div class="alert alert-info"><b>#NHSCrisis?  A quick fact check about the National Healthcare Service, NHS.</b><br/> 
Is the number of nurses in the UK really strongly decreasing, as some media outlets suggest?  Or is the opposite true, as some people on Twitter have argued? In this post I am taking a quick look at the number of employees in British healthcare system, from 2009 to 2017. <br/>
Since 2014, the number of nurses indeed goes down for most subgroups. Most notably "outgoing" groups of nurses show marked decline. The number of "adult nurses" has stayed rather constant. 
This loss in specialty nurses is compensated by an increase in the number of ambulance drivers and &apos;allied health professions&apos;. See plots at bottom of page. 
</div>')
```

#### A short story about reproducibility 

 This blogpost was inspired by a [lengthy twitter thread](https://twitter.com/Jeremy_Hunt/status/953688801825755136) on the 'NHSCrisis', the understaffed healthcare sector in the UK. Someone posted an infographics on a timeline with [official statistical data](https://digital.nhs.uk/article/4304/Workforce) from the British healthcare authorities. [He has argued](https://twitter.com/maxwellflitton) that there was a lot of secular (= cyclical but noisy) variation in the employee data, and that general the trend was pointing upwards, not downwards. Is this true?
 
 Are there *really* more and more nurses employed in the UK? Is it oversimplified to call this an '#NHSCrisis' on social media? I've decided to check this. 
 
The infographics the user posted was from a media company. (I cannot really include it here because it is probably copyrighted, and since late January, 2017 the user has protected his Twitter account, so I cannot link to the infographics in context). What struck my eye was that the graph ended early, in mid-2017. However it is now January 2018, and a full half year of data was missing. 

What would the graph look like when *I* tried to reproduce it from data?

A lot of underlying data are online at the above-mentioned NHS data portal. Also the this very newest data from the [second half of 2017 is online](https://digital.nhs.uk/catalogue/PUB30189) since this week, late January 2018. 

The time series we're interested in can be calculated from the timeseries data **&apos;NHS Workforce Statistics - staff excluding medical:&apos;**. What we need is an aggregated number: 

+ the total number of  nurses +
+ the number of 'community visitors' 

I think 'community visitors' are nurses that provide caregiving by visiting sick people at home - the immobile, the very sick or the elderly.

### Preprocessing 

After downloading the data, the analysis is straightforward


```{r libs}
library(lubridate)
library(tidyverse)
library(gridExtra)
theme_set(theme_bw()) 
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set2")
library(kableExtra)
```

Read in data files:

```{r read_csv}
# where data files are actually stored (~ 1 GB)
my_read_csv <- function(fn) read_csv(sprintf("../../../R_stuff/R_diverse/medicine_nhs_uk/%s", fn))

nhs_raw_2010 <- my_read_csv("NHS Workforce Statistics - Sep09-Dec10, staff excluding medical.csv")
nhs_raw_2011 <- my_read_csv("NHS Workforce Statistics - Jan11-Dec11, staff excluding medical.csv")
nhs_raw_2012 <- my_read_csv("NHS Workforce Statistics - Jan12-Dec12, staff excluding medical.csv")
nhs_raw_2013 <- my_read_csv("NHS Workforce Statistics - Jan13-Dec13, staff excluding medical.csv")
nhs_raw_2014 <- my_read_csv("NHS Workforce Statistics - Jan14-Dec14, staff excluding medical.csv")
nhs_raw_2015 <- my_read_csv("NHS Workforce Statistics - Jan15-Mar16, staff excluding medical.csv")
nhs_raw_2016 <- my_read_csv("NHS Workforce Statistics - September 2016, staff excluding medical.csv")
nhs_raw_2017 <- my_read_csv("NHS Workforce Statistics - September 2017, staff excluding medical.csv")

```

#### Raw data - combined 

Make a single large data frame
```{r bind_rows}
nhs_raw <- bind_rows(nhs_raw_2010, nhs_raw_2011, nhs_raw_2012, 
                     nhs_raw_2013, nhs_raw_2014, nhs_raw_2015, 
                     nhs_raw_2016, nhs_raw_2017)

```

This data frame has `r nrow(nhs_raw)` and `r ncol(nhs_raw)` columns. 

### Comparison 2009-2017

Count By Staff Group:

```{r count1}
mi <- min(nhs_raw$Date) 
mx <- max(nhs_raw$Date)
nhs_summ1 <- nhs_raw %>% 
        select(Date, `Main Staff Group`, `Total FTE`) %>%
        filter(Date %in% c(mi,mx)) %>% 
        group_by(Date, `Main Staff Group`) %>% 
        summarise(cnt = sum(`Total FTE`)) %>% 
        spread(Date, cnt)   

nhs_summ1 %>% 
        kable(align="lrr", row.names = 1, 
              caption = "Employee numbers at aggregation level 1,\
              at begin and end of the timeline") %>% 
        kableExtra::kable_styling(bootstrap_options = 
                                          c("striped", "hover"))

```

Staff numbers seem to have increased among all subgroups,
except for "038_NHS infastrucure support". 

This level of aggregation is useful, but it is too coarse. We only look at groups up and including to ID 019_Healthcare_scientists.

Calculate the totals at the appropriate aggregation level, *Staff Group 2* :

```{r count2}

nhs_summ2 <- nhs_raw %>% 
        select(Date, `Staff Group 2`, `Total FTE`) %>% 
        group_by(Date, `Staff Group 2`) %>% 
        summarise(cnt = sum(`Total FTE`))

```



#### Aggregation level 2

```{r eval=TRUE}
nhs_summ2 %>% spread(Date, cnt) %>% 
        select(`Staff Group 2`, 2, 82) %>% head(20) %>% 
        kable(caption="Aggregation Level 2 shows the nurse subgroups:",
              row.names=1)

```

For plotting, we are only interested in the groups including *019_Healthcare scientists*.

#### Plot: employee-count over time

These time series plots show of the employee count over time, for each nurse subgroup. I prefer to show them side-by-side here just to save space. Showing them one under the other would make the blogpost too long. 

Code to generate this plot is also shown.

```{r}
plot1 <- nhs_summ2 %>%
        filter(str_detect(`Staff Group 2` , "^00") == TRUE) %>% 
ggplot(aes(Date, cnt, color=`Staff Group 2`)) +
        geom_point(size=0.2) +
        geom_line() +
        ylab("Number of People") +        
        theme(legend.position = "none") +
        facet_wrap(~`Staff Group 2`, ncol = 1, scales = "free_y")

plot2 <- nhs_summ2 %>%
        filter(str_detect(`Staff Group 2` , "^01") == TRUE) %>% 
ggplot(aes(Date, cnt, color=`Staff Group 2`)) +
        geom_point(size=0.2) +
        geom_line() +
        theme(legend.position = "none") +
        ylab("Number of People") +
        facet_wrap(~`Staff Group 2`, ncol = 1, scales = "free_y")


```


```{r fig.height=30, echo=FALSE, fig.cap="Temporal evolution of 'non-medical UK nursing staff'"}

pl <- list(plot1, plot2)
ml <- marrangeGrob(pl, nrow=1, ncol=2, top = "Temporal evolution of ‘non-medical UK nursing staff’")
## non-interactive use, multipage pdf
#ggsave("multipage.pdf", ml, height = 30, units = "cm")
ml
```

### Further Ideas

- add data from Germany for comparison 
- calculate linear models, determine subgroups of strongest decline/growth.
- Add link / reference to MicroTrends book ( I want to reproduce more analyses as in the book.)
