---
title: Global interest in Bitcoin over Time
author: Knut Behrends
date: '2018-01-20'
draft: false
slug: bitcoin-interest
categories: ['personal-blog']
tags: ['code', 'finance', 'time-series']
banner: img/post/thumb/btc-1--2013-2018-150.png
output:
  html_document:
    toc: false
---

```{r intro,echo=FALSE}
htmltools::HTML('<div class="alert alert-info"><b>Bitcoin:</b>
Visualizing (1) Wikipedia Pageviews, (2) Google Searches and (3) Exchange Rates from 2015 to early 2018. 
Simple timeseries plots (created from Web-API data) show that these three are strongly correlated, except when they aren\'t.
</div>')
```

In this post I'll analyse global interest in [Bitcoin](https://en.wikipedia.org/wiki/Bitcoin) and [Ethereum](https://en.wikipedia.org/wiki/Ethereum) over time, according to accesses to their Wikipedia articles and as Google Searches. 

The end of the year 2017 saw the "Bitcoin [Bubble](https://www.bloomberg.com/news/audio/2017-03-31/this-is-how-you-know-when-the-stock-market-is-in-a-bubble)", a period of one month where 1 Bitcoin was traded for up to $20000 on most exchanges. I am not invested in Bitcoin, and never was substantially, but cryptocurrencies and the blockchain technology still interest me.

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, 
               echo = TRUE, tidy = FALSE, height = 3, width = 5,
               fig.width = 9)
options(digits = 3, tibble.print_min = 6)
```

### Preprocessing 

Preprocessing entails:

- Define necessary R packages
- Fetching the data

At this time, no user registration process is required to get to the data. 
No API keys and authentication steps are necessary to call the Web-APIs mentioned below. 

#### Load  R packages

```{r libs, cache=FALSE}

library(pageviews) # access Wikipedia
library(Quandl)    # provider of financial data
library(lubridate)
library(tidyverse) 
library(gtrendsR)  # access Google trends API
# ggplot defaults
theme_set(theme_bw()) 
scale_colour_discrete <- function(...) scale_color_brewer(palette="Set1")

```

#### Define a wrapper function to access a Wikipedia API

 
The function contains quite a few  of hardwired parameters. The [`pageviews`](https://github.com/ironholds/pageviews) package fetches the data. 
The function could be made a lot more flexible by adding more arguments, but that is not necessary for this blogpost.

The search function only works for terms for which an article  exists *with the same title* in both the German and the English Wikipedia.

Luckily "Bitcoin" is such a term.  

```{r func}
# Get the pageviews for "Search term"
# pageview data is available since 2015
startdate <- ymd("20150101")
enddate <- now()

wikipedia_de_vs_en <- function(term = "Bitcoin") {
  # call English wikipedia        
  pageviews <- article_pageviews(
    project = "en.wikipedia",
    article = term,
    user_type = "user",
    start = startdate,
    end = enddate
  )
  # call German wikipedia
  pageviews2 <- article_pageviews(
    project = "de.wikipedia",
    article = term,
    user_type = "user",
    start = startdate,
    end = enddate
  )

  # combine to data frame
  pageviews.df <- bind_rows(pageviews, pageviews2) %>%
    mutate(language = case_when(
      language == "en" ~ "English Wikipedia",
      language == "de" ~ "German Wikipedia"
    ))

  #plot it with 2 panels
  ggplot(pageviews.df, aes(date, views, color = language)) +
    geom_line() +
    xlab("") + ylab("Views per Day") +
    theme(legend.position = "none") +
    facet_wrap(~ language, nrow = 2, scales = "free_y") +
    ggtitle(
      sprintf("Daily Pageviews of '%s' in the English and German Wikipedias", term),
      subtitle = "Human Users only"
    ) 
}
```

### Data Visualization

(If you want to follow along, or verify the data: An interactive frontend that does the same is available at this url: [https://tools.wmflabs.org](https://tools.wmflabs.org/pageviews/?project=en.wikipedia.org&platform=all-access&agent=user&range=latest-20&pages=Bitcoin) 
)

Get some data for the Wikipedia article on Bitcoin:

```{r btc, echo=FALSE}
wikipedia_de_vs_en("Bitcoin") 
```

Get some data for the article on Ethereum, another popular cryptocurrency (at this time).

```{r eth, echo=FALSE}
wikipedia_de_vs_en("Ethereum") 
```

These plots show a lot of things. 

- Unsurprisingly, the interest (measured in pageviews) in the English version of the Wikipedia articles is consistently higher than interest in the German version. About 5 times as high.
- Accesses to the Wikipedia articles peaked quite a few times even before the 2017 craze started. (I haven't looked it up what these events were. Ransomware spreads, perhaps?). 
- Interest in Ethereum seems to be disproportionally high in Germany: English pageviews were "only" 3 times as high as German pageviews (usually it's 5 times as high).  
- Strong peaks in Ethereum-article pageviews  often occured independently from Bitcoin-Article pageviews, for both language articles.
- (TBC)

There is another, alternative data source: Google Trends API, here accessed with functions from the [gtrendsR](https://github.com/PMassicotte/gtrendsR) package. 
The API provides aggregated and rescaled data on how many people have performed Google searches for any term.

```{r goog, echo=FALSE}
btcgoog <- here::here("data/bitcoin_googletrends.RDS")
if(file.exists(btcgoog)){
        bitcoin_trends <- readRDS(file = btcgoog)
} else{
        bitcoin_trends <- gtrends("Bitcoin", geo = "", time = "today+5-y", 
                                  gprop = c("web"), category = 0, hl = "en-US")
        saveRDS(bitcoin_trends, btcgoog)
}
```

Show a plot for comparison:

```{r gogl}

bitcoin_trends$interest_over_time %>%
  filter(date >= as.Date("2015-01-01")) %>%
  ggplot(aes(date, hits)) +
  geom_line() +
  theme(legend.position = "none") +
  xlab("") + ylab("Interest") +
  ggtitle(
    sprintf("Google Trends: Worlwide Interest over Time in '%s'", "Bitcoin"),
    subtitle = "(Google internal counters)"
  )
```

In order to make the Google-Trends-API plot comparable to the other plots above,
I have restricted the data to the interval from 2015 to 2018. Thus I have left out a small but notable spike in the gloabl interest in bitcoin that happended in  2014. 

#### Another obligatory comparison

Last but not least:  the exchange rate for Bitcoin, here given in Euro. Data are from [Bitcoin.de](https://www.bitcoin.de), aggregated by [Quandl](https://docs.quandl.com/docs/time-series).

```{r quandl}
btc <- Quandl(code = "BCHARTS/BTCDEEUR", type = "xts")
autoplot(btc$Close["2015-01-01/"]) +
        ylab("Bitcoin/EUR") + xlab("Year") +
        ggtitle("Bitcoin Exchange Rate 2015-2018",
                subtitle="'Close Price' in EUR, according to the Bitcoin.de trading platform. Data from Quandl.com.")

```

What many people talk about is the 'volatility' of Bitcoin. 
Rarely have I seen someone visualizing this, so here is another plot showing the daily change in the 'closing price' *(do these trading platforms really close?)* for 1 Bitcoin, in Euro:

```{r btcdiff}
autoplot(diff(btc$Close["2015-01-01/"])) +
        ylab("Daily Price Change, in EUR/day ") + xlab("Year") +
        ggtitle("Volatility of Bitcoin 'Close Price' 2015-2018",
                subtitle="in EUR, according to the Bitcoin.de trading platform. Data from Quandl.com.")
```

I've never seen any financial time series like that.