---
title: 'Constructing a tiny social-network visualization from the #Rstats hashtag'
author: Knut Behrends
date: '2018-01-01'
slug: constructing-a-tiny-social-network-visualization-from-the-rstats-hashtag
categories: [blog-post]
tags: [fun, rstats]
banner: img/post/thumb/jq-med.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# (Under Construction - Checking if 3-D plot below works )

## Constructing a tiny social-network visualization from the #Rstats hashtag

In 2016, I was harvesting tweets from the Twitter streaming API. I've used the `#rstats` hashtag as the search filter. This hashtag is more descriptive than the `#R` hashtag which may not even be a valid filter expression because it is so short - only two characters.

During a day or so I gathered about 100 tweets, encoded in JSON format.  
Each tweet contains about 2500 characters. This might surprise you, because a tweet used to be 140 characters in length, until Twitter increased it to 280 characters in 2017. But a tweet contains a lot of metadata.


In a different blogpost ('[Learning jq](/post/2017/12/learning-jq)')  I've explained how to extract specific fields from the nested structure. See example 4.

This text file can be read in as a data frame:

```{r message=FALSE, warning=FALSE}
options(knitr.table.format = "html") 
library(here)
library(assertthat)
library(tidyverse)
library(igraph)
datadir <- "data/twitter"
see_if(is.readable(datadir))

infile1 <- here::here(datadir, "rstats_network.txt")
mentions <- read_csv(infile1)
# remove leading "_" from column names
colnames(mentions) <- gsub("^_", "", colnames(mentions), perl = TRUE)
```

Convert it to an igraph object, a network
```{r df2igr}
users <- mentions %>% 
        select(name) %>% 
        dplyr::union(mentions %>% select(mentions_name) %>% rename(name ="mentions_name"))


gr <- mentions %>% 
        select(name, mentions_name) %>%
        unique() %>% 
        filter(name != mentions_name) %>% 
        graph_from_data_frame(directed = TRUE, vertices = users)

gr <- set_vertex_attr(gr, "color", value = "dodgerblue")

# Redraw the graph and make the vertex size 1
library(threejs)

ec <- as.numeric(eigen_centrality(gr)$vector)

# Create new vector 'v' that is equal to the square-root of 'ec' multiplied by 5
v <- 2 * sqrt(ec)

# Plot threejs plot of graph setting vertex size to v
graphjs(gr, vertex.size = v, vertex.label = TRUE, main="#Rstats tweet network, 2016")

```

```{r moreplots, eval=FALSE, echo=FALSE}
g.b <- betweenness(gr, directed = TRUE)
#Show histogram of vertex betweenness
hist(g.b, breaks = 80)


plot(gr,
     vertex.label = NA,
     vertex.label.color = "black", 
     edge.color = 'gray77',
     vertex.size = 0,
     edge.arrow.size = 0.1,
     layout = layout_nicely(gr))

# Create plot with vertex size determined by betweenness score
plot(gr, 
#     vertex.label = NA,
     edge.color = 'black',
     vertex.size = sqrt(g.b)+1,
     edge.arrow.size = 0.05,
     layout = layout_nicely(gr))
```

Show some details of the network

```{r}
incident(gr, 'Hadley Wickham', mode = c("all")) %>%  head(20)
```

