---
title: Drawing a network with "ggraph" and "tidygraph"
draft: true
author: Knut Behrends
date: '2018-01-16'
slug: drawing-a-figshare-network-with-ggraph-and-tidygraph
categories:
  - R
  - networks
tags:
  - 'rstats'
  - 'tidyverse'
---

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
               echo = TRUE, tidy = FALSE, height = 3, width = 5,
               fig.width = 9)
options(digits = 3, tibble.print_min = 6)

```

```{r intro,echo=FALSE}
htmltools::HTML('<div class="alert alert-success alert-info alert-warning alert-danger"><b>NOTE</b>
This blogpost is inspired by this one: <a href="https://nxskok.github.io/blog/2017/12/30/drawing-a-network-with-ggraph-and-tidygraph/">
Drawing a network with "ggraph" and "tidygraph" by Ken Butler
</a><br>
</div>')
```

I'm trying to learn the differences between using the *igraph* library directly and the new packages *ggraph* and *tidygraph*.


## Introduction



[Thomas Lin Pedersen](https://github.com/thomasp85) is the author of a
lot of packages, including two that deal with graphs (in the sense of
networks), `tidygraph` for storing and handling graphs, and `ggraph` for
drawing them, `ggplot`-style. I have a feeling I will be spending a lot
of time with `tidygraph`, but in this post, I get my feet wet reading in
and storing a graph and then making a picture of it.

Packages:

```{r libs}
options(knitr.table.format = "html") 
options(fig.width = 9) 
library(igraph)
library(httr)
library(tidyverse)
library(tidygraph)
library(ggraph)
theme_set(theme_bw())
```


## Example: the Figshare categories


## Reading in the data and storing as a graph

I had to think about how I wanted to represent the network. If these had
been plane flights, I would have had a starting point and an ending
point for each one. The input file (part) looks like this:


```{r}

categories_file <- "data/network/figshare-categories.json"

# if(! file.exists(categories_file)) {
        categories <- httr::GET("http://api.figshare.com/v2/categories") %>%
                content(as="text") %>%
                jsonlite::fromJSON(simplifyDataFrame = TRUE) %>% 
                  as_tibble() 
#         write(categories, file = categories_file)
# } 
# categories <- jsonlite::fromJSON(categories_file), 
#                   simplifyDataFrame = TRUE) %>% 
#                   as_tibble()

```



```
#head --lines=21 figshare-categories.json | tail -n+2
 < figshare-categories.json jq . | head --lines=21 - | tail -n+2
```


To read this in, I treat it as a three-column json-File. 
Quick look at the data:

```{r}
categories %>% head(n=10)
```

Quick visualization how category IDs were assigned by Figshare Designers

```{r vis1}
p1 <- ggplot(categories, aes(id, I(1:nrow(categories)))) +
        geom_point(size=0.1) +
        xlab("Category ID") + ylab("Count") +
        ggtitle("Figshare Category IDs", subtitle="Gaps indicate that a few category-IDs have been deleted or renamed")
p2 <-  ggplot(categories, aes(parent_id)) +geom_histogram(binwidth = 5) +
        xlab("Category Parent IDs") + ylab("Count") +
        ggtitle("Figshare Parent Category IDs", subtitle="Gaps indicate that parent-category-IDs have small values or few distinct values > 1000")
gridExtra::grid.arrange(p1,p2, nrow=2, ncol=1)

```

Which category-ids are *not* assigned?

```{r}
unused_ids <- setdiff(1:max(categories$id), unique(categories$id))

ggplot(data.frame(pos = 1:length(unused_ids), id=unused_ids), aes(pos, id)) +
        geom_point(size=0.1) +
        scale_x_continuous(limits=c(0,50)) +
        ggtitle("Unused Category IDs", 
                subtitle = "Roughly every 3rd number is not assigned to a category IDs")
```



`tidygraph` has a function for turning a data frame structured like
this, with edges as rows, into an actual tidy graph:

```{r graph}

categories_g <- categories %>%
        select(id , parent_id) %>% 
        as_tbl_graph()

categories_g

```

How many attribute names are set?

```{r}
categories_g %>%
        vertex_attr() %>% 
        names()
```

Only one attribute "name" is set.

Which edges have the parent_id 48?

```{r eval=TRUE}

igraph::E(categories_g)[[inc('48')]]  %>%
        head(10)


# similar call
# incident(categories_g, '48', mode = c("in"))
```

There are `r length(igraph::E(categories_g)[[inc('48')]]  )` such edges.


```{r}

categories_g <- set_edge_attr(categories_g,
                              "subject",
                              value=categories$title)
categories_g <- set_edge_attr(categories_g,
                              "id",
                              value=categories$id)
categories_g <- set_edge_attr(categories_g,
                              "parent_id",
                              value=categories$parent_id)
```

Now, how to assign names to the category ids.

The nodes and edges are shown separately, with the nodes numbered (those
numbers are used in the edges).

## Extracting the nodes and edges

`tidygraph` uses the same pipeline ideas as the `tidyverse`, but extra
care is required because a tidy graph object has both nodes and edges in
it. An extra verb `activate` lets you specify which one you would like
to be "primary" (but you can also access the other one). Piping the
output of `activate` into `as_tibble` lets you save the nodes or edges
as a data frame, separately:

```{r}

nodes = categories_g %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(node_number=row_number()) %>%
  print(n=10)
```

```{r}
edges = categories_g %>%
  activate(edges) %>%
  as_tibble() %>% print(n=10)
```

I saved the node numbers as well as names into the nodes data frame.
This lets me look up the *names* of the nodes referred to in the edges,
to convince myself that I have something sensible:

```{r}
edges %>%
  left_join(nodes,by=c("from"="node_number")) %>%
  left_join(nodes,by=c("to"="node_number")) %>%
  print(n=20)
```

This is what I started with.

## Drawing the graph


Thomas Lin Pedersen also wrote the package `ggraph` which allows
`ggplot`-style plotting of graphs. For our example, this was my ultimate
aim: I wanted to see how well the connection information enabled us to
reproduce the map above.

Graphs can be plotted in many different ways. For this graph, I know
there is a representation in two dimensions, since the stations are
scattered across the 2D-space of West categorieshire. In `ggraph`, this
is controlled by the `layout` parameter of the `ggraph` command.  The
Kamada-Kawai method tries to plot neighbouring nodes a constant distance
apart, and chooses their locations as if they are connected by
unit-length springs. I find this works well here. In addition, I want to
draw the edges (as lines), draw the nodes (as points), and label the
nodes by which station they are. These are accomplished by
`geom_edge_link`, `geom_node_point` and `geom_node_text`. Additional
notes: the `alpha=0.3` draws the edges partly transparent, so that the
station names stand out; the use of `repel=T` "repels" the station names
away from the points they represent so that you can see them (in the
manner of `geom_text_repel`):

```{r}
ggraph(categories_g,layout="kk")+
  geom_edge_link(colour="blue", alpha=0.1)+
  geom_node_point()

 # geom_node_text(aes(label=id),size=2,repel=T)
```

This seems to do a nice job of reproducing the topology in a clear
fashion with no overlaps where there shouldn't be any. (I tried this
using the separate stations, Westgate and Kirkgate, in Wakefield, and
the display was a lot more confusing.) It doesn't get compass directions
right (it seems to have west and east reversed), but we had no right to
expect that it would.

Another plausible layout is the one called `lgl`. How does that work
here? We leave everything else the same:

```{r}
ggraph(categories_g,layout="lgl")+
  geom_edge_link(colour="blue",alpha=0.1)+
  geom_node_point()
#  geom_node_text(aes(label=name),size=2,repel=T)
```


```{r}
g.outd <- degree(categories_g, mode = c("in"))
#names(g.outd) <- edge_attr(categories_g, "subject")
#View a summary of out-degree
table(g.outd)
```

