---
title: Global interest in Bitcoin over Time
author: Knut Behrends
date: '2018-01-20'
draft: false
slug: bitcoin-interest
categories: ['personal-blog']
tags: ['code', 'finance', 'time-series']
banner: img/post/thumb/btc-1--2013-2018-150.png
output:
  html_document:
    toc: false
---

```{r intro,echo=FALSE}
htmltools::HTML('<div class="alert alert-info"><b>Abstract:</b>
Bitcoin. Visualizing (1) Wikipedia Pageviews, (2) Google Searches and (3) Exchange Rates from 2015 to early 2018. 
Simple timeseries plots (created from Web-API data) show that these three are strongly correlated.
</div>')
```

In this post I'll analyse global interest in Bitcoin and Ethereum over time, according to accesses to Wikipedia articles and as Google Searches. At this time, no API keys and authentication are necessary.

The end of the year 2017 saw the "Bitcoin Bubble", a period of one month where 1 Bitcoin was traded for up to $20000 on most exchanges. I am not invested in Bitcoin, and never was substantially, but cryptocurrencies and the blockchain technology still interest me.

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
               echo = TRUE, tidy = FALSE, height = 3, width = 5,
               fig.width = 9)
options(digits = 3, tibble.print_min = 6)
```

### Preprocessing 

Preprocessing entails:

- Define necessary R packages
- Fetching the data


#### Load  R packages

```{r libs, cache=FALSE}

library(pageviews) # access Wikipedia
library(Quandl)    # provider of financial data
library(lubridate)
library(tidyverse) 
library(gtrendsR)  # access Google trends API
theme_set(theme_bw()) # ggplot default

```

#### Define a wrapper function to access a Wikipedia API

 
The function contains quite a few  of hardwired parameters. The [`pageviews`](https://github.com/ironholds/pageviews) package fetches the data. 
The function could bemade a lot more flexible by adding more parameters but that is also not necessary for this blogpost.

The search function only works for terms that for which an article in both wikipedias exists. 

Luckily "Bitcoin" is such a term.

```{r func}
# Get the pageviews for "Search term"
# pageview data is available since 2015
startdate <- ymd("20150101")
enddate <- now()

wikipedia_de_vs_en <- function(term = "Bitcoin") {
  # call English wikipedia        
  pageviews <- article_pageviews(
    project = "en.wikipedia",
    article = term,
    user_type = "user",
    start = startdate,
    end = enddate
  )
  # call German wikipedia
  pageviews2 <- article_pageviews(
    project = "de.wikipedia",
    article = term,
    user_type = "user",
    start = startdate,
    end = enddate
  )

  # combine to data frame
  pageviews.df <- bind_rows(pageviews, pageviews2) %>%
    mutate(language = case_when(
      language == "en" ~ "English Wikipedia",
      language == "de" ~ "German Wikipedia"
    ))

  #plot it with 2 panels
  ggplot(pageviews.df, aes(date, views, color = language)) +
    geom_line() +
    xlab("") + ylab("Views per Day") +
    theme(legend.position = "none") +
    facet_wrap(~ language, nrow = 2, scales = "free_y") +
    ggtitle(
      sprintf("Daily Pageviews of '%s' in the English and German Wikipedias", term),
      subtitle = "Human Users only"
    ) 
}
```

Get some data for Bitcoin

```{r btc, echo=FALSE}
wikipedia_de_vs_en("Bitcoin") 
```

Get some data for Ethereum, another popular cryptocurrency (at this time).

```{r eth, echo=FALSE}
wikipedia_de_vs_en("Ethereum") 
```

There is another, alternative data source: Google Trends API, here accessed with functions from the [gtrendsR](https://github.com/PMassicotte/gtrendsR) package. 
The API provides aggregated and rescaled data on how many people have performed Google searches for this term.

```{r goog, echo=FALSE}
btcgoog <- here::here("data/bitcoin_googletrends.RDS")
if(file.exists(btcgoog)){
        bitcoin_trends <- readRDS(file = btcgoog)
} else{
        bitcoin_trends <- gtrends("Bitcoin", geo = "", time = "today+5-y", 
                                  gprop = c("web"), category = 0, hl = "en-US")
        saveRDS(bitcoin_trends, btcgoog)
}
```

Show a plot for comparison:

```{r gogl}

bitcoin_trends$interest_over_time %>%
  filter(date > as.Date("2015-01-01")) %>%
  ggplot(aes(date, hits)) +
  geom_line() +
  theme(legend.position = "none") +
  xlab("") + ylab("Interest") +
  ggtitle(
    sprintf("Google Trends: Worlwide Interest over Time in '%s'", "Bitcoin"),
    subtitle = "(Google internal counters)"
  )
```

#### Another obligatory comparison

Last but not least:  the exchange rate for Bitcoin, here given in Euro. 

```{r quandl}

btc <- Quandl(code = "BCHARTS/BTCDEEUR", type = "xts")
autoplot(btc$Close["2015-01-01/"]) +
        ylab("Bitcoin/EUR") + xlab("Year") +
        ggtitle("Bitcoin Exchange Rate 2015-2018",
                subtitle="in EUR, according to the Kraken.com trading platform")

```

